Capturing underlying musical structure: an audio-based approach to the \textit{Ursatz} using stems and deep learning

### 2023-02-09 introduction/introduction.tex
Updates from Overleaf

Music Structure Analysis (MSA) is the field of study that aims to understand music's underlying organization and structure. This can include analyzing elements such as melody, harmony, rhythm, and timbre, as well as the relationships and interactions between these elements. MSA aims to identify patterns and regularities in music and develop computational models that can replicate and generate musical structures.

The fundamental concept behind MSA is the ability to divide a song into distinct, non-overlapping segments, assigning a label to each component that defines its type. This segmentation and labeling process aims to capture a human's perception or analysis of the song.

@@

\section{Topic overview}

Music Structure Analysis (MSA) has a long history, dating back to music theory and composition studies. In the 18th and 19th centuries, music theorists such as Rameau~\cite{christensen2004rameau} and Schenker~\cite{schenkerdocumentsonline} developed analytical frameworks for understanding the structure of music in terms of harmony, melody, and form.

@@

As a music lover and connoisseur, I am interested in gaining a deeper understanding of the fundamental building of musical composition and the various techniques used to create musical works. I am deeply interested in learning how to retrieve embedded and encoded information within the music. The study of music structure analysis provides a means to systematically analyze and understand musical pieces. It has traditionally been approached through music theory, formal musical notation, musical analysis techniques, and machine learning, which I plan to blend.

With the advent of artificial intelligence and machine learning, the field of music structure analysis has expanded to encompass the use of musically motivated neural network architectures. These networks are designed to incorporate musical concepts and perceptual approaches and use this knowledge to analyze and understand the structure of musical pieces in a way that goes beyond traditional music analysis techniques.

As a professional music transcriber, you are interested in understanding how deep neural networks (DNNs) filter musical features. In other words, I am eager to learn how DNNs listen~\cite{7500246} to music.

@@

>>>>

Music\footnote{The term "music" used in this context refers specifically to Western tonal music tradition. It is assumed that the musical structures and elements discussed are based on this tradition and may not apply to other musical styles or cultures.} Structure Analysis (MSA) is the field of study that aims to understand music's underlying organization and structure.

@@

@@

As a music lover and connoisseur, I am interested in gaining a deeper understanding of the fundamental building of musical composition, the various techniques used to create musical works, and the underlying relationship among them. I am deeply interested in learning how to retrieve embedded information within music. The study of music structure analysis provides a means to systematically analyze and understand musical pieces. It has traditionally been approached through music theory, formal musical notation, musical analysis techniques, and machine learning, which I plan to blend.

With the advent of artificial intelligence and machine learning, Music Information Retrieval (MIR) has expanded to encompass musically motivated neural network architectures. These networks are designed to incorporate musical concepts and perceptual approaches and use this knowledge to analyze and understand the structure of musical pieces in a way that goes beyond traditional music analysis techniques.

As a professional music transcriber, I am interested in understanding how deep neural networks (DNNs) filter musical features. In other words, I am eager to learn how DNNs listen~\cite{7500246} to music.

@@

In this report, we have organized the information into several sections and sub-sections to provide a clear and comprehensive overview of the topic. The structure of the report is as follows:

\begin{itemize}
\item Introduction
  \begin{itemize}
  \item Purpose
  \item Methodology
  \item Scope
  \end{itemize}
\item Main Body
  \begin{itemize}
  \item Section 1
  \item Section 2
  \item Section 3
  \end{itemize}
\item Conclusion
  \begin{itemize}
  \item Summary
  \item Recommendations
  \end{itemize}
\item References
  \begin{itemize}
  \item Sources Cited
  \end{itemize}
\end{itemize}

### 2023-02-09 music analysis/MSA.tex
Updates from Overleaf

MSA encompasses various aspects of music analysis, including algorithms, current practices, and trends in technology. This field aims to comprehensively understand music structure by applying musicological theories and principles and music signal processing algorithms. MSA considers the interplay between technical and musical aspects, focusing on improving our understanding of music signals and the music itself.

In the following section, we will examine the state of MSA, including its challenges and potential for future advancements. Topics such as data representation, computational complexity, and evaluation metrics will be discussed to provide a nuanced summary of the field. Additionally, we will consider the impact of new trends and technological advancements on MSA and the directions in which the field is likely to evolve. Through this examination, we aim to provide a thorough understanding of the technical and musical aspects of MSA and its challenges.

~\cite{Nieto2020}~\cite{Chaki2021}

@@

The musical structure analysis (MSA) challenges the definition due to subjectivity, ambiguity, and data scarcity. A flat structural analysis of an audio signal consists of a set of contiguous, non-overlapping, and labeled time intervals that span the entire duration of the audio signal. Given a musical recording of T audio samples, a flat segmentation is defined by a set of segment boundaries B, a set of k unique labels Y, and a mapping of segment starting points to labels S. This leads to the label assignment L, which assigns a label to each time point. The segment boundaries are usually sampled at a rate of 10 Hz to achieve a balance between resolution and computational efficiency.

As an art form, music is open to interpretation and can evoke different emotions and meanings in different listeners. This subjectivity makes it difficult to objectively analyze the structure of a piece of music, as different annotators may place boundaries between segments differently based on their perception and interpretation of the music. It can also have multiple structures, including hierarchical structures, that make it difficult to untangle the truth.

\subsection{Segmentation principles}
The principles of musical structure analysis (MSA) were initially defined as homogeneity, novelty, and repetition, with the addition of regularity later on. To analyze the structure of a musical recording, a self-similarity matrix (SSM) is often used to visualize the degree of similarity between audio frames. The SSM results in a square matrix with the highest degree of similarity on the diagonal.

@@

\section{Theoretical approach}
\subsection{Music form}

Music form refers to the structural organization of a musical composition or performance. This can include elements such as the arrangement of musical units (such as rhythm, melody, and harmony) that exhibit repetition or variation, the arrangement of instruments, or the orchestration of a symphonic piece. These elements interact to shape the composition and create a meaningful musical experience for the listener. The organizational elements can be further broken down into smaller units called phrases, which express a musical idea but lack sufficient weight to stand alone. The form of a composition unfolds over time through the expansion and development of these ideas. In tonal harmony, form is primarily articulated through cadences, phrases, and periods.

@@

@@

Schenkerian analysis uses specialized musical notation and is either presented in a generative manner, starting from the Ursatz and ending at the score or in a reductive manner, starting from the score and reducing it to its fundamental structure. The Ursatz is presented in an arrhythmic graph, and rhythmic signs display hierarchical relationships between pitch events.

>>>>

MSA encompasses various aspects of music analysis, including algorithms, current practices, and trends in technology. This field aims to comprehensively understand music structure by applying musicological theories, principles, and signal processing algorithms. MSA considers the interplay between technical and musical aspects, focusing on improving our understanding of music signals and the music itself.

In the following section, we will examine the state of MSA, including its challenges and potential for future advancements. Topics such as data representation, computational complexity, and evaluation metrics will be discussed to provide a nuanced summary of the field. Additionally, we will consider the impact of new trends and technological advancements on MSA and the directions in which the field is likely to evolve. Through this examination, we aim to provide an overview of MSA's technical and musical aspects and its challenges~\cite{Nieto2020}~\cite{Chaki2021}.

\section{Concept definition}

Musical structure and form refer to the organization and arrangement of musical elements in a piece of music. They describe how the music is put together and how it unfolds over time.

Musical structure hierarchically arranges various elements of music, such as melody, harmony, rhythm, timbre, and texture\footnote{Musical texture refers to the arrangement and combination of musical elements, including melody, harmony, rhythm, timbre, and dynamics, in a piece of music. Different textures, such as monophonic, homophonic, polyphonic, and heterophonic, can affect a piece's overall character and mood. Understanding musical texture is vital for music analysis and appreciation.}. This hierarchy can take different forms, such as melody being the most critical element, harmony playing a supportive role, or rhythm being the driving force behind the music. The structure of a piece of music is often characterized by its elements' repetition, development, and variation.

Music form, however, refers to the larger-scale organization of musical structure over time. It describes how music sections are repeated, varied, and put together to create a complete piece. Music form can be thought of as the plan or design that underlies a piece of music. Many forms are used in music, such as binary, ternary, rondo, sonata, and theme and variations.

Musicologists and theorists have developed various analytical tools and methods for understanding musical structure and form, including Schenkerian analysis, set theory, and formal function analysis. These tools and methods help uncover the underlying musical relationships and patterns in a piece of music and provide a framework for understanding how it is put together and how it develops over time.

@@

The musical structure analysis (MSA) challenges the definition due to subjectivity, ambiguity\footnote{Perception and interpretation of musical structure can vary widely between individuals and depend on factors such as personal background, cultural context, and individual preferences. This is due to the complex interplay between individual differences in perception and interpretation and how the auditory system processes sound. These factors highlight the importance of considering individual and cultural context in analyzing and interpreting musical structure.}, and data scarcity. A flat structural analysis of an audio signal consists of a set of contiguous, non-overlapping, and labeled time intervals that span the entire duration of the audio signal. Given a musical recording of T audio samples, a flat segmentation is defined by a set of segment boundaries B, a set of k unique labels Y, and a mapping of segment starting points to labels S. This leads to the label assignment L, which assigns a label to each time point. The segment boundaries are usually sampled at a rate of 10 Hz to achieve a balance between resolution and computational efficiency.

As an art form, music is open to interpretation and can evoke different emotions and meanings in different listeners. This subjectivity makes it difficult to objectively analyze the structure of a piece of music, as different annotators may place boundaries between segments differently based on their perception and interpretation of the music. It can also have multiple structures, including hierarchical structures ~\cite{hiererchical}, that makes it difficult to untangle the underlying structure.

\subsection{Segmentation principles}
The main principles of musical structure analysis (MSA) were initially defined as homogeneity, novelty, and repetition, with the addition of regularity later on.

@@

\section{Common methods}

To analyze the structure of a musical recording, a self-similarity matrix (SSM) is often used to visualize the degree of similarity between audio frames. The SSM results in a square matrix with the highest degree of similarity on the diagonal.

It can be computed by representing the audio signal as a spectrogram (or any other suitable representation) and then calculating the similarity between all pairs of time frames in the spectrogram.

One common approach is to use a dot product, where the similarity between two-time frames is equal to the dot product of their vectors. The dot product of two vectors x and y is given by:

\input{music analysis/Equations/SSM equation}

The resulting matrix will be symmetric, with the diagonal elements representing the similarity between a time frame and itself and the off-diagonal elements representing the similarity between two different time frames.

It's also common to normalize the similarity values by dividing each entry by the Euclidean norm of the compared vectors to ensure that the self-similarity matrix has values between -1 and 1.

Given a self-similarity matrix S, the novelty curve can be computed by taking the row-wise sum of the self-similarity matrix and normalizing the values. The novelty curve measures the deviation from the average similarity across time frames and can be used for various audio analysis tasks such as beat tracking or segmentation.

\input{music analysis/Equations/Novelty equation}

This representation is used for various tasks, such as identifying repeating patterns in the audio, detecting structural changes in the audio, or for music classification.

\section{Musicological approach}
\section{Topic overview}

MSA has a long history, dating back to music theory and composition studies. In the 18th and 19th centuries, music theorists such as Rameau~\cite{christensen2004rameau} and Schenker~\cite{schenkerdocumentsonline} developed analytical frameworks for understanding the structure of music in terms of harmony, melody, and form.

@@

The fundamental structure, also known as \textit{Ursatz}, represents the underlying structure of a tonal work at the most remote or background level, in the most abstract form. It comprises an essential elaboration of the tonic triad, including the entire line and accompanying bass arpeggiation. The fundamental structure, like the fundamental line, takes one of three forms, depending on which pitch of the tonic triad serves as the primary tone. For example, the key of C major would represent the fundamental structure by the fundamental descending line from scale degree 3.

The \textit{Urlinie}, or fundamental line, unfolds the basic triad and presents tonality on horizontal paths. The tonal system also flows into this to bring a sense of purposeful order to chords by selecting harmonic degrees. Voice leading serves as the mediator between the horizontal formulation of tonality presented by the Urlinie and the vertical formulation presented by the harmonic degrees.

The upper voice of the fundamental structure, represented by the whole line, utilizes a descending direction. In contrast, the lower voice, the bass arpeggiation through the fifth, takes an ascending direction (as shown in figure 1). The combination of the fundamental line and bass arpeggiation constitutes a unity, and neither can exist alone; it is only through their unification in a contrapuntal structure that they produce art.

@@

Schenkerian analysis uses specialized musical notation and is either presented in a generative manner, starting from the Ursatz and ending at the score or in a reductive manner, starting from the score and reducing it to its fundamental structure. The Ursatz is presented in an arrhythmic graph, and rhythmic signs display hierarchical relationships between events.\footnote{In Schenkerian analysis, structural levels refer to the representation of a piece of music at different levels of abstraction. These levels typically include the foreground, middle ground, and background. Schenker posits that musical form, as represented through these structural levels, can be understood as an energy transformation, specifically the transformation of forces that flow from the background to the foreground through the various levels.}
